---
title: "01_Import_and_MAE"
author: "Dupret Florent"
date: "`r Sys.Date()`"
output:
  html_document: 
    df_print: kable
    theme: yeti
    highlight: tango
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.show='hold', cache = FALSE, results = 'hold', warning = FALSE, message = FALSE, error = FALSE)
#Cleaning rstudio
rm(list = ls())
#Libraries required
require(ggplot2)
require(tidyverse)
require(FactoMineR)
require(readr)
require(SummarizedExperiment)
require(MultiAssayExperiment)
```

# Data preparation

Data preparation was done in R studio (version XXX) and involves importing the raw datafiles, selecting the information of interest and organizing it into a central structured object.

The first step was to import the datafiles \textit{mae\_7\_with\_MC.Rds},  \textit{Gates\_metabolomics\_batch\_accounting.csv},  \textit{gates\_metabolomics\_022420\_stanford.csv} and  \textit{gates\_metabolomics\_061219\_UMD.csv} and to rename them \emph{res}, \emph{batch}, \emph{stan} and \emph{umd} respectively. Since the data originates from a larger research project than my master’s thesis, we already know a lot about it and what it contains. Therefore, with my promoter, we were able to identify relevant pieces of data for my project, omitting to consider other pieces. In that sense, some of the following procedures might be a based on given instructions or prior information and need no justification. The \emph{res} and \emph{batch} file both contain sample relevant data which would be combined into the same column data object. \emph{res} also contained a SummarizedExperiment object with the RNA count data for the vaginal microbiota composition of both UMD and Stanford cohort members. Finally, objects \emph{stan} and \emph{umd} contained the metabolomic data of Stanford and UMD respectively.

After importing the files, the batch data was rectified: the sample ID column was shortened, and a new column combining the cohort and the batch components into a single string of characters was created. This was the first element of the column data. The second element was contained in the \emph{res} \emph{MultiAssayExperiment} (MAE) object. From this file, the following variables of interest to this project were selected: \emph{SampleID, Subject, Status, Cohort, Race, BMI, Age, GestationalAge\_days, Bleeding, PH, cycle\_nb, cycle\_nb\_m, cycle\_length, cycleday, MC\_ok} and \emph{Reprod\_status}. After combining the variables with the batch data into a new dataframe called \emph{coldata}, this newly created object was separated according to cohort: one column data objet for Stanford, called \emph{MB\_stan\_col} and one for UMD, called \emph{MB\_umd\_col}.

The metabolomics datasets were tackled next: the row relevant metadata was isolated from the main metabolomic peak intensity data and renamed \emph{MB\_stan\_row} and \emph{MB\_umd\_row} accordingly. Two extra rows from the Stanford metabolomics data were discarded as they did not contain any data (completely empty rows). These main data tables, renamed \emph{MB\_stan\_assay} and \emph{MB\_umd\_assay}, were isolated from the rest. Then, the assay data, row and column metadata were collected into a \emph{SummarizedExperiment} object called \emph{MB\_stan} and \emph{MB\_umd} accordingly. The row names and column names were also corrected for ease of manipulation.

As a second-to-last step, the gene count data was prepared: only the variables of interest were selected in the column data. It was then separated according to cohort allocation into the object \emph{VM\_stan} and \emph{VM\_umd}.

Finally, two MultiAssayExperiment objects were created: one for Stanford cohort and the other for UMD. They both comprised the microbiota composition data (gene counts - \emph{VM\_stan} or \emph{VM\_umd}) and metabolite data (peak intensities - \emph{MB\_stan} or \emph{MB\_umd}) with their respective row and column metadata element. These MAE objects will be the central data source for the subsequent data preprocessing steps and analyses.

```{r}
res <- readRDS(file= "../03_Data/mae_7_with_MC.Rds")
#Extracting the VMRC dataset
VMRC <- res@ExperimentList$VM16S_VMRC_ASVseq
```

```{r}
#Preparation of the "BATCH" dataset
#Import batch dataset
batch <- read_csv("../03_Data/Gates_metabolomics_batch_accounting.csv")
#Rectify SampleID column and combining Cohort+Batch
batch <- batch %>%
  mutate(SampleID = UID %>%
           str_remove_all(" [0-9]*$") %>%
           str_replace_all(" ","_"),
         Batch = Metabolomics_batch,
         Cluster = paste(Cohort, Metabolomics_batch)) %>%
  select(SampleID, Batch, Cluster)
#Ordering batch according to sampleId
batch <- arrange(batch, SampleID)
```

```{r, eval = FALSE, echo = FALSE}
#METABOLITE PREPARATION
#Import MB datasets
stan <- read_csv(file = "../03_Data/gates_metabolomics_022420_stanford.csv")
umd <- read_csv(file = "../03_Data/gates_metabolomics_061219_UMD.csv")
#Identifying dissimilarities between umd and stan datasets
odd <- stan$BIOCHEMICAL[stan$BIOCHEMICAL != umd$BIOCHEMICAL]
#Correcting differences
summary(t(stan[odd,6:ncol(stan)]))
stan <- stan[-which(stan$BIOCHEMICAL %in% odd),]
#Creating respective SummarizedExperiment objects
MB_stan <- SummarizedExperiment(assay = stan[6:ncol(stan)], rowData = stan[1:5])
MB_umd <- SummarizedExperiment(assay = umd[6:ncol(umd)], rowData = umd[1:5])
#Adapting Stanford column names
MB_stan <- MB_stan[,sort(MB_stan@colData@rownames)]
#Adapting UMD column names
colnames(MB_umd) <- colnames(MB_umd) %>% str_remove_all("[0-9]*$") %>% str_trim(side = "right") %>% str_replace_all(" ","_")
MB_umd <- MB_umd[,sort(MB_umd@colData@rownames)]
#Adapting Stanford row names
rownames(MB_stan) <- stan$BIOCHEMICAL
#Adapting UMD row names
rownames(MB_umd) <- umd$BIOCHEMICAL
```
```{r}
#METABOLITE PREPARATION
#Import MB datasets
stan <- read_csv(file = "../03_Data/gates_metabolomics_022420_stanford.csv")
umd <- read_csv(file = "../03_Data/gates_metabolomics_061219_UMD.csv")

#Identifying dissimilarities between umd and stan datasets
odd <- stan$BIOCHEMICAL[stan$BIOCHEMICAL != umd$BIOCHEMICAL]
#Correcting differences
summary(t(stan[odd,6:ncol(stan)]))
stan <- stan[-which(stan$BIOCHEMICAL %in% odd),]

#Creating rowData based on the first 5 columns of Metabolomic dataset
MB_stan_row <- stan[1:5]
MB_umd_row <- umd[1:5]

#Creating assay based on the columns of Metabolomic dataset (200 columns)
MB_stan_assay <- stan[6:ncol(stan)]
MB_umd_assay <- umd[6:ncol(umd)]

#Adapting Stanford column names
MB_stan_assay <- MB_stan_assay[,sort(colnames(MB_stan_assay))]
#Adapting UMD column names
colnames(MB_umd_assay) <- colnames(MB_umd_assay) %>% str_remove_all("[0-9]*$") %>% str_trim(side = "right") %>% str_replace_all(" ","_")
MB_umd_assay <- MB_umd_assay[,sort(colnames(MB_umd_assay))]

#Adapting Stanford row names
rownames(MB_stan_assay) <- stan$BIOCHEMICAL
#Adapting UMD row names
rownames(MB_umd_assay) <- umd$BIOCHEMICAL
```

```{r, eval = FALSE, echo = FALSE}
#COLUMN DATA PREPARATION
#Create 
coldata <- colData(res)
#Selecting columns
ColsOfInterest <- c("SampleID", "Subject", "Status", "Cohort", "Race", "BMI", "Age", "GestationalAge_days", "Bleeding", "PH", "cycle_nb", "cycle_nb_m", "cycle_length", "cycleday", "MC_ok", "Reprod_status")
coldata <- coldata[,ColsOfInterest]
#Filtering rows
coldata <- coldata[coldata$SampleID %in% batch$SampleID,]
#Sorting coldata by SampleID
coldata <- arrange(data.frame(coldata), SampleID)
#Combining batch data and coldata
coldata <- left_join(coldata, batch, by = "SampleID")
#Separating coldata according to Pregnant (Stanford) and Non-pregnant (UMD) status
coldata.stan <- coldata[coldata$Status == "Pregnant",]
coldata.umd <- coldata[coldata$Status == "Non-pregnant",]
```
```{r}
#COLUMN DATA PREPARATION
#Create 
coldata <- colData(res)
#Selecting columns
ColsOfInterest <- c("SampleID", "Subject", "Status", "Cohort", "Race", "BMI", "Age", "GestationalAge_days", "Bleeding", "PH", "cycle_nb", "cycle_nb_m", "cycle_length", "cycleday", "MC_ok", "Reprod_status")
coldata <- coldata[,ColsOfInterest]

#Filtering rows
coldata <- coldata[coldata$SampleID %in% batch$SampleID,]

#Sorting coldata by SampleID
coldata <- arrange(data.frame(coldata), SampleID)

#Combining batch data and coldata
coldata <- left_join(coldata, batch, by = "SampleID")

#Setting rownames of coldata as SampleID
rownames(coldata) <- coldata$SampleID

#Separating coldata according to Pregnant (Stanford) and Non-pregnant (UMD) status
MB_stan_col <- coldata[coldata$Status == "Pregnant",]
MB_umd_col <- coldata[coldata$Status == "Non-pregnant",]
```

```{r}
#METABOLOMIC SUMMARIZEDEXPERIMENT OBJECT
MB_stan <- SummarizedExperiment(assay = MB_stan_assay,
                                rowData = MB_stan_row,
                                colData = MB_stan_col)
MB_umd <- SummarizedExperiment(assay = MB_umd_assay,
                               rowData = MB_umd_row,
                               colData = MB_umd_col)
```

```{r, eval = FALSE, echo = FALSE}
#Selecting relevant variables in VMRC
colData(VMRC) <- colData(VMRC)[c("SampleID", "BarcodeSequence", "Platform", "NumberInRun", "AmpDNAconc", "SampleSum")]
#Missing colnames
mis.colnames <- coldata[!coldata$SampleID %in% colnames(VMRC),]$SampleID
#Creating Gap Filler
GapFiller <- matrix(nrow = nrow(VMRC), ncol = 7,
                    dimnames = list(rownames(VMRC),mis.colnames))
GapFiller.coldata <- matrix(nrow = 7, ncol = ncol(colData(VMRC)),
                            dimnames = list(mis.colnames ,colnames(colData(VMRC))))
#Combining column data
coldata.VMRC <- rbind(colData(VMRC), data.frame(GapFiller.coldata))
#Combining assay data
assay.VMRC <- cbind(assay(VMRC), GapFiller)
#Recreating the VMRC SummarizedExperiment from scratch
VMRC <- SummarizedExperiment(assays = assay.VMRC,
                             rowData = rowData(VMRC),
                             colData = coldata.VMRC)
#Sorting SE according to colnames (sample ID)
VMRC <- VMRC[,sort(colnames(VMRC))]
#Separating VMRC according to Stanford and UMD patient ID
VM_stan <- VMRC[,VMRC@colData@rownames %in% colnames(MB_stan)]
VM_umd <- VMRC[,VMRC@colData@rownames %in% colnames(MB_umd)]
```
```{r}
#Selecting relevant variables in VMRC
VMRC_col <- colData(VMRC)[c("SampleID", "BarcodeSequence", "Platform", "NumberInRun", "AmpDNAconc", "SampleSum")]
#Adding batch variable to column data
VMRC_col <- left_join(x = data.frame(VMRC_col), y = batch, by = "SampleID")

#Assay data
VMRC_assay <- assay(VMRC)
#Row data
VMRC_row <- rowData(VMRC)

#Recreating the VMRC SummarizedExperiment from scratch
VMRC <- SummarizedExperiment(assays = VMRC_assay,
                             rowData = VMRC_row,
                             colData = VMRC_col)
#Sorting SE according to colnames (sample ID)
VMRC <- VMRC[ ,sort(colnames(VMRC))]
#Separating VMRC according to Stanford and UMD patient ID
VM_stan <- VMRC[,VMRC@colData@rownames %in% colnames(MB_stan)]
VM_umd <- VMRC[,VMRC@colData@rownames %in% colnames(MB_umd)]
```

```{r}
#Creating assay list
assay_list_stan <- list(MB = MB_stan,
                        VM = VM_stan)
assay_list_umd <- list(MB = MB_umd,
                       VM = VM_umd)
#Creating MultiAssayExperiment
STANFORD <- MultiAssayExperiment(experiments = ExperimentList(assay_list_stan))
UMD <- MultiAssayExperiment(experiments = ExperimentList(assay_list_umd))
```

The end-products of this preparation section are summarized below:

```{r, results='markup'}
#Checking results
STANFORD
UMD
```


* `MB` = pre-processed peak intensities of metabolites samples measured using semi-targeted mass spectrometry (MS). This SummarizedExperiment has contains an assay with `r ncol(STANFORD@ExperimentList$MB)` columns and `r nrow(STANFORD@ExperimentList$MB)` rows in Stanford, and `r ncol(UMD@ExperimentList$MB)` columns and `r nrow(UMD@ExperimentList$MB)` rows in UMD There are approximately 5 samples from 40 pregnant women in Stanford and 5 samples from 40 non-pregnant women in UMD. 

  - Its `rowData` has `r nrow(rowData(STANFORD[["MB"]]))` columns (for both Stanford and Umd) expressing the name of the biochemicals / metabolites, the super-pathway to which they are part of, the sub-pathway to which they are part of, the compound's mass and its KEGG identification code.

  - The `colData` with batch related information (batch, sample ID, cohort) and clinical data for each cohort member like gestational age, age, BMI, vaginal pH, cycle day, etc., for a total of `r ncol(colData(STANFORD[["MB"]]))` variables in Stanford and `r ncol(colData(UMD[["MB"]]))` variables in UMD.

* `VM` = 16S rRNA sequence counts of `r nrow(STANFORD@ExperimentList$VM)` (`r nrow(UMD@ExperimentList$VM)`) identified bacterial strands from the vaginal ecosystem of Stanford patients (UMD patients respectively). These counts come from `r ncol(STANFORD@ExperimentList$VM)` (`r ncol(UMD@ExperimentList$VM)`) samples of the same 40 pregnant women in Stanford (or the same 40 non-pregnant women in UMD).

  - This object also contains row data of dimension `r dim(STANFORD@ExperimentList$VM@elementMetadata)` in Stanford and `r dim(UMD@ExperimentList$VM@elementMetadata)` in UMD
  - It also contains its specific column data of dimension `r dim(STANFORD@ExperimentList$VM@colData)` (`r dim(UMD@ExperimentList$VM@colData)`) with information specific to the RNASeq measurements (like the library size) and batch allocation

It is easy to understand the dimensions of the metabolite (MB) object or the vaginal microbiota (VM) are the same in both Stanford and UMD.

```{r}
#Saving results
saveRDS(STANFORD, file = "../05_Results/STANFORD_mae.Rds")
saveRDS(UMD, file = "../05_Results/UMD_mae.Rds")
```
